Implications of afforestation policies for global markets

### 2024-02-26 paper.Rmd
Update paper.Rmd

More recent modeling attempts of land-use change have begun to move away from comparative static analysis. For example, the dynamic GTAP model, supported by the GTAP-AEZ database, considers land market effects, which were identified as significant in driving results by Stevenson et al. [@stevenson2013], and has been utilized to evaluate the impacts of crop intensification on land use. It, however, has been critiqued, as global aggregates may mask localized shifts that can have implications for ecologically significant areas [@byerlee2014]. In addition, in the standard GTAP model, it is difficult to model a sector individually. There are 68 market sectors covered in the latest version of the GTAP database. The GTAP model's general equilibrium nature means that all these markets are cleared with market-clearing prices. As the model applies nested CES production functions to all sectors and differentiates them by parameters only, to model one sector, in this case forestry, differently from the rest would require a major redesign of the model. While not impossible, this project will attempt to get around these concerns and assess the impact of various likely land-use scenarios by melding GTAP with the Global Timber Model (GTM), which is a dynamic optimization model that has been specifically designed to analyze the relationship among land rents, forested land cover, and carbon sequestration through forests and thus is better able to examine the impact of afforestation or avoided deforestation on the carbon sink [@sohngen2001]. Joining these two models and passing outputs back and forth between the models should enable any limitations of either model to be minimized and more robust results, therefore, to be attained.

@@

Establishing a connection between the standard GTAP model and the GTM model was a two-step process. First, we identified the set of common variables that are included in both models and their closures. That is, we selected which variables would be determined endogenously by one model and held as exogenous in the other model. In our case, we modeled the return on land using the GTAP model and fed those results into the GTM model as exogenous. Then, the amount of land available for crop and livestock production was endogenously determined by the GTM model and fed into the GTAP model as an exogenous input. In the second step, we established the connection between the models on the software level. This step required that the GTAP model, which is written in GEMPACK, be able to handle with the inputs from the GTM model, which is written in GAMS, and submit the outputs back to the GTM model. As GEMPACK and GAMS were developed long before software integration became important or even possible, we had to create a set of user functions in R that executed both models, collected their outputs, and created parameter files automatically.

@@

# Scenarios

>>>>

More recent modeling attempts of land-use change have begun to move away from comparative static analysis. For example, the dynamic GTAP model, supported by the GTAP-AEZ database, considers land market effects, which were identified as significant in driving results by Stevenson et al. [@stevenson2013], and has been utilized to evaluate the impacts of crop intensification on land use. It, however, has been critiqued, as global aggregates may mask localized shifts that can have implications for ecologically significant areas [@byerlee2014]. In addition, in the standard GTAP model, it is difficult to model a sector individually. There are 68 market sectors covered in the latest version of the GTAP database. The GTAP model's general equilibrium nature means that all these markets are cleared with market-clearing prices. As the model applies nested CES production functions to all sectors and differentiates them by parameters only, to model one sector, in this case forestry, differently from the rest would require a major redesign of the model. While not impossible, this project will attempt to get around these concerns and assess the impact of various likely land-use scenarios by melding GTAP with the Global Timber Model (GTM), which is a dynamic optimization model that has been specifically designed to analyze the relationship among land rents, forested land cover, and carbon sequestration through forests and thus is better able to examine the impact of afforestation or avoided deforestation on the carbon sink [@sohngen2001]. The latest versions of both models were used. For the GTAP model, this meant using the 11b data which has 2017 as a base year. This data was then updated to the base year (2023) using data from the Centre for Prospective Studies and International Information (CEPII) to account for the projected changes in gross domestic product (GDP), population, and productivity for each country in the world.

@@

By joining these two models and passing outputs back and forth between the models, any limitations of either model should be minimized and more robust results, therefore, should be attained. Establishing a connection between the standard GTAP model and the GTM model required  a two-step process. First, we identified the set of common variables that are included in both models and their closures. That is, we selected which variables would be determined endogenously by one model and held as exogenous in the other model. In our case, we modeled the return on land using the GTAP model and fed those results into the GTM model as exogenous. Then, the amount of land available for crop and livestock production was endogenously determined by the GTM model and fed into the GTAP model as an exogenous input. In the second step, we established the connection between the models on the software level. This step required that the GTAP model, which is written in GEMPACK, be able to handle with the inputs from the GTM model, which is written in GAMS, and submit the outputs back to the GTM model. As GEMPACK and GAMS were developed long before software integration became important or even possible, we had to create a set of user functions in R that executed both models, collected their outputs, and created parameter files automatically.

@@

# Modeling the impact of possible measures to mitigate deforestation/promote afforestation

To understand the impact of potential regional and global afforestation actions, an important first step was creating a realistic baseline scenario that depicts the likely changes between over the decade under study: 2023-2033. This was done by taking into account key drivers of change like growth in population and GDP. Without accounting for this, results would likely have been biased as they would have failed to account for the increasing pressure to convert forests into agricultural land. Data for this were drawn from sources such as CEPII and the United Nations.

When examining the likely impacts of the baseline changes alone, it is estimated that by 2033, significant deforestation would have occured in many regions of the world. For example, in Brazil...
